---
title: "NADP_trend_analysis"
author: "Bella Oleksy"
date: "2024-11-19"
output: html_document
---


# Some notes at the top about data processing

<b>Interpolation method--</b>
-  Here, I am linearlly interpolated for periods of missing data with gaps no more than 2 weeks apart. I am using the NO3, NH4, Total IN, and precip values from  Bret A. Schichtel et al. 2019 Sci Tot Env.

<b>Year vs. water year--</b>
Aggregating data on an annual basis instead of WY. Just flagging that in case we switch to WY later.

```{r setup, include=FALSE, echo=FALSE}
#source scripts
source("scripts/00_functions.R")
source("scripts/00_libraries.R")
source("scripts/01.1_LochO_data_munging.R")
source("scripts/01.2_Bear_Lake_SNOTEL.R")
source("scripts/01.3_NADP_CO98.R")
knitr::opts_chunk$set(fig.width=6, fig.height=8)


```

## Missing data:
Data are missing throughout, unsurprisingly, but there are a few time periods that stick out, like 1986.
I will exclude them for the initial trend analysis because they are biased low.
```{r, echo=FALSE}
vis_miss(NADP%>%
           select(waterYear, weekofyear, subppt_mm, ppt_mm_bret, Ca_mgL, Mg_mgL, K_mgL, NH4_mgL, NO3_mgL, SO4_mgL, Na_mgL,
                  NO3_mgL_bret, NH4_mgL_bret, IN_mgL_bret) %>%
           filter(waterYear=="1986"))

```

```{r, echo=FALSE, include=FALSE}
#Visualize missing data with naniar package
vis_dat(NADP)
vis_miss(NADP%>%
           select(waterYear, weekofyear, subppt_mm, ppt_mm_bret, Ca_mgL, Mg_mgL, K_mgL, NH4_mgL, NO3_mgL, SO4_mgL, Na_mgL,
                  NO3_mgL_bret, NH4_mgL_bret, IN_mgL_bret))

#We are missing a ton of data in 1986 so the interpolations will still underestimate.
#Remove for trend analyses?

ggplot(NADP, 
       aes(x = weekofyear, 
           y = NO3_mgL)) + 
  geom_miss_point() +
  facet_wrap(~waterYear)+
  labs(title="Raw NADP data")

ggplot(NADP, 
       aes(x = weekofyear, 
           y = NO3_mgL_bret)) + 
  geom_miss_point() +
  facet_wrap(~waterYear)+
  labs(title="Bret's NADP data")

# NADP_trim <- NADP %>%
#   select(waterYear, weekofyear, ppt_mm_bret, Ca_mgL, Mg_mgL, K_mgL, NH4_mgL_bret, NO3_mgL_bret, IN_mgL_bret, SO4_mgL, Na_mgL) %>%
#   mutate(cations_mgL = Ca_mgL + Mg_mgL + Na_mgL + K_mgL) %>%
#   select(-c(Ca_mgL, Mg_mgL, Na_mgL, K_mgL)) %>%
#   pivot_longer(-c(1:3),
#                names_to = "deposition_name",
#                values_to = "deposition_value_mg_L") %>%
#   mutate(rainwater_m3 = (ppt_mm_bret/1000)*loch_ws_size_m2, # over the entire watershed that week
#          deposition_value_kg_m3 = deposition_value_mg_L / 1e+3, # Convert  concentration from mg/L to kg/m³
#          deposition_kg = rainwater_m3 * deposition_value_kg_m3, #kg across the time period (a week in this case)
#          deposition_kg_per_ha = deposition_kg/660) 



```


### Test for annual deposition trends

Demo of what's happening under the hood -- run the M-K test, and if significant extract the Sens slope and intercept.

```{r, echo=FALSE}

#Here is what's happening under the hood
N_dep_annual <- annual_dep %>%
  filter(deposition_name=="Inorganic N") %>%
  filter(waterYear > 1984 & waterYear < 2024)
# test the data for normality
hist(N_dep_annual$deposition_kg_per_ha_impute, breaks = 30)
# but let's use a Shapiro test to check for real
shapiro.test(N_dep_annual$deposition_kg_per_ha_impute)  
#P-value is below .05, therefore we reject the null hypothesis of a normal distrubution.

# We can turned to the Mann-Kendall trend test, which is non-parametric, not requiring a normal dist
# this only tells us if there is a significant trend, not the estimated slope
mk_results <-trend::mk.test(N_dep_annual$deposition_kg_per_ha_impute)
print(mk_results) # significant

# for a slope estimate we uses the Sen's slope function
sens_result <-trend::sens.slope(N_dep_annual$deposition_kg_per_ha_impute)
print(sens_result)
slope_est <-as.numeric(sens_result[[1]]) # slope estimate

# the trend package only gives us slope, not a y-int estimate for plotting
# we'll use the zyp package for that
y_int_results <-zyp::zyp.sen(deposition_kg_per_ha_impute ~ waterYear, N_dep_annual) # y comes first!!
print(y_int_results) # inspect
y_int <-y_int_results$coefficients[[1]] # pull out y-int estimate for ploting

ggplot(N_dep_annual, aes(y = deposition_kg_per_ha_impute, x = waterYear)) +
  geom_point(col = "red", size=5)+
  # use info from our MK and Sen's test to graph the trend line
  geom_abline(intercept = y_int, slope = slope_est, color="blue", size=1.2) +
  labs(title="Annual N deposition MK Trend", 
       y="Inorganic N (kg/ha)", x="Water Year")+
  theme_few(base_size=36)


## Relationship between annual dep and ppt?
head(annual_dep)


```


```{r, echo=FALSE}
#The sens slope p-value matches M-K so I built a workshop that just uses the sens slope and then the zyp package to get a y-intercept

##ANNUAL

annual_dep_trends_nested <- annual_dep %>%
  rename(value=deposition_kg_per_ha_impute) %>%
  filter(waterYear > 1984 & waterYear < 2023) %>%
  # filter(!waterYear==1986) %>% #Lots of missing data for c("SO4-S","cations")
  group_by(deposition_name) %>%
  nest() %>%
  mutate(
    sens = map(data, map_sens),
    sens_sum = map(sens, broom::glance),
    slope = map(sens, sens_slope),
    zyp_mod = map(data, map_zyp),
    intercept = map(zyp_mod, sens_intercept)
  )


## Un-nest 
annual_dep_trends_unnested = unnest(annual_dep_trends_nested, c(sens_sum, slope, deposition_name)) %>%
  mutate(
    trend = case_when(
      p.value <= 0.05 & slope >= 0 ~ 'increasing',
      p.value <= 0.05 & slope <= 0 ~ 'decreasing',
      p.value > 0.05 ~ 'no trend'
    ),
    trend = factor(trend,
                       levels = c('no trend',
                                  'increasing',
                                  'decreasing'))
  )

#Extract all covariates for modeling...
annual_dep_trends <- annual_dep_trends_unnested %>%
  select(deposition_name, slope, p.value, intercept, trend) %>%
  unnest(c(slope)) %>%
  mutate(time_frame = "annual")%>%
  mutate(intercept = as.numeric(intercept))



##MONTHLY

monthly_dep_trends_nested <- monthly_dep %>%
  rename(value=deposition_kg_per_ha_impute) %>%
  filter(waterYear > 1984 & waterYear < 2023) %>%
  # filter(!waterYear==1986) %>% #Lots of missing data for c("SO4-S","cations")
  group_by(deposition_name, month) %>%
  nest() %>%
  mutate(
    sens = map(data, map_sens),
    sens_sum = map(sens, broom::glance),
    slope = map(sens, sens_slope),
    zyp_mod = map(data, map_zyp),
    intercept = map(zyp_mod, sens_intercept)
  )


## Un-nest 
monthly_dep_trends_unnested = unnest(monthly_dep_trends_nested, c(sens_sum, slope, deposition_name, month)) %>%
  mutate(
    trend = case_when(
      p.value <= 0.05 & slope >= 0 ~ 'increasing',
      p.value <= 0.05 & slope <= 0 ~ 'decreasing',
      p.value > 0.05 ~ 'no trend'
    ),
    trend = factor(trend,
                       levels = c('no trend',
                                  'increasing',
                                  'decreasing'))
  )

#Extract all covariates for modeling...
monthly_dep_trends <- monthly_dep_trends_unnested %>%
  select(deposition_name, month, slope, p.value, trend, intercept) %>%
  unnest(c(slope)) %>%
  rename(time_frame=month)%>%
  mutate(intercept = as.numeric(intercept))


##SUMMER

summer_dep_trends_nested <- summer_dep %>%
  rename(value=deposition_kg_per_ha_impute) %>%
  filter(waterYear > 1984 & waterYear < 2023) %>%
  # filter(!waterYear==1986) %>% #Lots of missing data for c("SO4-S","cations")
  group_by(deposition_name) %>%
  nest() %>%
  mutate(
    sens = map(data, map_sens),
    sens_sum = map(sens, broom::glance),
    slope = map(sens, sens_slope),
    zyp_mod = map(data, map_zyp),
    intercept = map(zyp_mod, sens_intercept)
  )


## Un-nest 
summer_dep_trends_unnested = unnest(summer_dep_trends_nested, c(sens_sum, slope, deposition_name)) %>%
  mutate(
    trend = case_when(
      p.value <= 0.05 & slope >= 0 ~ 'increasing',
      p.value <= 0.05 & slope <= 0 ~ 'decreasing',
      p.value > 0.05 ~ 'no trend'
    ),
    trend = factor(trend,
                       levels = c('no trend',
                                  'increasing',
                                  'decreasing'))
  )

#Extract all covariates for modeling...
summer_dep_trends <- summer_dep_trends_unnested %>%
  select(deposition_name, slope, p.value, trend, intercept) %>%
  unnest(c(slope)) %>%
  mutate(time_frame = "summer")%>%
  mutate(intercept = as.numeric(intercept))


all_dep_trends <- bind_rows(annual_dep_trends,
                            summer_dep_trends,
                            monthly_dep_trends) %>%
  mutate(slope = round(slope, 3),
         p.value = round(p.value,4),
         p.value = case_when(p.value < 0.001 ~ "<0.001",
                             TRUE ~ as.character(p.value))) 

all_dep_trends_hux <- 
  hux(all_dep_trends) %>% 
  # add_colnames() %>% 
  set_bold(row = 1, col = everywhere, value = TRUE) %>% 
  set_all_borders(TRUE) 

theme_plain(all_dep_trends_hux)
# quick_docx(all_dep_trends_hux, file = 'figures/Table1_deposition_trends.docx')

```

### Annual deposition 1984-2022

There is only a trend line on the plot if p < 0.05.

```{r, echo=FALSE, warning=FALSE}

annual_dep %>%
  rename(value = deposition_kg_per_ha_impute) %>%
  filter(waterYear > 1984 & waterYear < 2023) %>%
  filter(!waterYear == 1986) %>%
  ggplot(aes(
    x = waterYear,
    y = value,
    fill = deposition_name,
    group = deposition_name
  )) +
  geom_point(size = 2, shape = 21) +
  facet_wrap( ~ deposition_name, scales = "free_y", ncol=2) +
  theme_few(base_size = 12) +
  geom_abline(
    data = annual_dep_trends %>% filter(!trend == "no trend"),
    aes(
      intercept = intercept,
      slope = slope,
      color = deposition_name
    )
  ) +
  labs(y = "Annual deposition (kg/ha/year)",
       x = "Year") +
  # geom_smooth(method = "gam") +
  scale_color_manual(values = chemCols,
                     "Fit") +
  scale_fill_manual(values = chemCols,
                    name = "Observed") +
  theme(legend.position = "none")
ggsave(
  "figures/deposition_trends_annual.png",
  dpi = 600,
  width = 4,
  height = 6,
  units = "in"
)
```

### Monthly deposition 1984-2022

There is only a trend line on the plot if p < 0.05.

```{r, echo=FALSE, warning=FALSE}

monthly_dep %>%
  rename(value=deposition_kg_per_ha_impute) %>%
  filter(waterYear > 1984 & waterYear < 2023) %>%
    filter(!waterYear==1986) %>% #Lots of missing data for c("SO4-S","cations")
  ggplot(aes(x=waterYear, y=value, fill=month, group=month))+
  geom_point(size=2, shape=21)+ 
  facet_wrap(~deposition_name, scales="free_y")+
  theme_few(base_size = 12)+
  geom_abline(data = monthly_dep_trends %>% rename(month=time_frame)  %>% filter(!trend=="no trend"), aes(intercept = intercept, slope = slope, color=month)) +
  labs(y="Deposition (kg/ha/year)",
       x="Year",
       title="Monthly") +
  scale_color_manual(values=monthCols,
                     "Fit")+
  scale_fill_manual(values=monthCols,
                    name="Observed")
ggsave("figures/deposition_trends_monthly.png", dpi=600, width=8, height=6, units="in")
```

### Summer deposition rates (1984-2022)

There is only a trend line on the plot if p < 0.05.

```{r, echo=FALSE, warning=FALSE}

summer_dep %>%
  rename(value=deposition_kg_per_ha_impute) %>%
  filter(waterYear > 1984 & waterYear < 2023) %>%
  filter(!waterYear==1986) %>% #Lots of missing data for c("SO4-S","cations")
  ggplot(aes(x=waterYear, y=value, fill=deposition_name, group=deposition_name))+
  geom_point(size=2, shape=21)+ 
  facet_wrap(~deposition_name, scales="free_y")+
  theme_few(base_size = 12)+
  geom_abline(data = summer_dep_trends %>% filter(!trend=="no trend"),
              aes(intercept = intercept, slope = slope, color=deposition_name)) +
  labs(y="Deposition (kg/ha/year)",
       x="Year",
       title="Summer") +
  scale_color_manual(values=chemCols,
                     "Fit")+
  scale_fill_manual(values=chemCols,
                    name="Observed")
ggsave("figures/deposition_trends_summer.png", dpi=600, width=8, height=6, units="in")
```
