---
title: "NADP_trend_analysis"
author: "Bella Oleksy"
date: "2024-11-19"
output: html_document
---


# Some notes at the top about data processing

<b>Interpolation method--</b>
- In the future we'll want to (I think) mirror the method by Bret A. Schichtel et al. 2019 Sci Tot Env. Here, I am linearlly interpolated for periods of missing data with gaps no more than 2 weeks apart 

<b>Year vs. water year--</b>
Aggregating data on an annual basis instead of WY. Just flagging that in case we switch to WY later.

```{r setup, include=FALSE, echo=FALSE}
#source scripts
source("scripts/00_functions.R")
source("scripts/00_libraries.R")
source("scripts/01.1_LochO_data_munging.R")
source("scripts/01.2_Bear_Lake_SNOTEL.R")
source(here("scripts/01.3_NADP_CO98.R"))
knitr::opts_chunk$set(fig.width=unit(18,"cm"), fig.height=unit(11,"cm"))
```

## Missing data:
Data are missing throughout, unsurprisingly, but there are a few time periods that stick out, like 1986.
I will exclude them for the initial trend analysis because they are biased low.
```{r, echo=FALSE}
vis_miss(NADP%>%
           select(waterYear, weekofyear, subppt_mm, ppt_mm_bret, Ca_mgL, Mg_mgL, K_mgL, NH4_mgL, NO3_mgL, SO4_mgL, Na_mgL,
                  NO3_mgL_bret, NH4_mgL_bret, IN_mgL_bret) %>%
           filter(waterYear=="1986"))

```

```{r, echo=FALSE, include=FALSE}
#Visualize missing data with naniar package
vis_dat(NADP)
vis_miss(NADP%>%
           select(waterYear, weekofyear, subppt_mm, ppt_mm_bret, Ca_mgL, Mg_mgL, K_mgL, NH4_mgL, NO3_mgL, SO4_mgL, Na_mgL,
                  NO3_mgL_bret, NH4_mgL_bret, IN_mgL_bret))

#We are missing a ton of data in 1986 so the interpolations will still underestimate.
#Remove for trend analyses?

ggplot(NADP, 
       aes(x = weekofyear, 
           y = NO3_mgL)) + 
  geom_miss_point() +
  facet_wrap(~waterYear)+
  labs(title="Raw NADP data")

ggplot(NADP, 
       aes(x = weekofyear, 
           y = NO3_mgL_bret)) + 
  geom_miss_point() +
  facet_wrap(~waterYear)+
  labs(title="Bret's NADP data")

# NADP_trim <- NADP %>%
#   select(waterYear, weekofyear, ppt_mm_bret, Ca_mgL, Mg_mgL, K_mgL, NH4_mgL_bret, NO3_mgL_bret, IN_mgL_bret, SO4_mgL, Na_mgL) %>%
#   mutate(cations_mgL = Ca_mgL + Mg_mgL + Na_mgL + K_mgL) %>%
#   select(-c(Ca_mgL, Mg_mgL, Na_mgL, K_mgL)) %>%
#   pivot_longer(-c(1:3),
#                names_to = "deposition_name",
#                values_to = "deposition_value_mg_L") %>%
#   mutate(rainwater_m3 = (ppt_mm_bret/1000)*loch_ws_size_m2, # over the entire watershed that week
#          deposition_value_kg_m3 = deposition_value_mg_L / 1e+3, # Convert  concentration from mg/L to kg/m³
#          deposition_kg = rainwater_m3 * deposition_value_kg_m3, #kg across the time period (a week in this case)
#          deposition_kg_per_ha = deposition_kg/660) 

#Annual dep
annual_dep <- NADP %>%
  select(waterYear, weekofyear, ppt_mm_bret, Ca_mgL, Mg_mgL, K_mgL, NH4_mgL_bret, NO3_mgL_bret, IN_mgL_bret, SO4_mgL, Na_mgL) %>%
  mutate(cations_mgL = Ca_mgL + Mg_mgL + Na_mgL + K_mgL) %>%
  select(-c(Ca_mgL, Mg_mgL, Na_mgL, K_mgL)) %>%
  pivot_longer(-c(1:3),
               names_to = "deposition_name",
               values_to = "deposition_value_mg_L") %>%
  # filter(deposition_name=="NO3_mgL") %>%
  distinct(deposition_name, waterYear, weekofyear, .keep_all = TRUE) %>%
  arrange(deposition_name,waterYear, weekofyear) %>%
  as_tsibble(., key = c(deposition_name,waterYear), index = weekofyear) %>% #time series tibble
  fill_gaps() %>%
  #impute missing data
  mutate(deposition_name = case_match(
    deposition_name,
    "cations_mgL" ~ "cations",
    # "inorganicN_mgL" ~ "inorganic N",
    "NH4_mgL_bret" ~ "NH4-N",
    "NO3_mgL_bret" ~ "NO3-N",
    "SO4_mgL" ~ "SO4-S",
    "IN_mgL_bret" ~ "Inorganic N",
    .default = deposition_name
  )) %>%
  mutate(deposition_value_mg_L = case_when(deposition_name == "NO3-N" ~ deposition_value_mg_L * 0.226,
                                deposition_name == "NH4-N" ~ deposition_value_mg_L * 0.777,
                                deposition_name == "SO4-S" ~ deposition_value_mg_L * 0.333,
                                TRUE ~ deposition_value_mg_L)) %>%
  mutate(ppt_mm_bret = imputeTS::na_interpolation(ppt_mm_bret, maxgap = 2),
         deposition_value_mg_L = imputeTS::na_interpolation(deposition_value_mg_L, maxgap = 2),
         rainwater_m3 = (ppt_mm_bret/1000)*loch_ws_size_m2, # over the entire watershed that week
         deposition_value_kg_m3 = deposition_value_mg_L / 1e+3, # Convert  concentration from mg/L to kg/m³
         deposition_kg = rainwater_m3 * deposition_value_kg_m3, #kg across the time period (a week in this case)
         deposition_kg_per_ha = deposition_kg/660) %>%
  as.data.frame() %>%
  group_by(waterYear, deposition_name) %>%
  summarize(deposition_kg_per_ha_impute = sum(deposition_kg_per_ha, na.rm=TRUE))

#Monthly dep (June-Aug)
monthly_dep <- NADP %>%
  select(waterYear, weekofyear, ppt_mm_bret, Ca_mgL, Mg_mgL, K_mgL, NH4_mgL_bret, NO3_mgL_bret, IN_mgL_bret, SO4_mgL, Na_mgL) %>%
  mutate(cations_mgL = Ca_mgL + Mg_mgL + Na_mgL + K_mgL) %>%
  select(-c(Ca_mgL, Mg_mgL, Na_mgL, K_mgL)) %>%
  pivot_longer(-c(1:3),
               names_to = "deposition_name",
               values_to = "deposition_value_mg_L") %>%
  # filter(deposition_name=="NO3_mgL") %>%
  distinct(deposition_name, waterYear, weekofyear, .keep_all = TRUE) %>%
  arrange(deposition_name,waterYear, weekofyear) %>%
  as_tsibble(., key = c(deposition_name,waterYear), index = weekofyear) %>% #time series tibble
  fill_gaps() %>%
  #impute missing data
  mutate(deposition_name = case_match(
    deposition_name,
    "cations_mgL" ~ "cations",
    # "inorganicN_mgL" ~ "inorganic N",
    "NH4_mgL_bret" ~ "NH4-N",
    "NO3_mgL_bret" ~ "NO3-N",
    "SO4_mgL" ~ "SO4-S",
    "IN_mgL_bret" ~ "Inorganic N",
    .default = deposition_name
  )) %>%
  mutate(deposition_value_mg_L = case_when(deposition_name == "NO3-N" ~ deposition_value_mg_L * 0.226,
                                deposition_name == "NH4-N" ~ deposition_value_mg_L * 0.777,
                                deposition_name == "SO4-S" ~ deposition_value_mg_L * 0.333,
                                TRUE ~ deposition_value_mg_L)) %>%
  mutate(ppt_mm_bret = imputeTS::na_interpolation(ppt_mm_bret, maxgap = 2),
         deposition_value_mg_L = imputeTS::na_interpolation(deposition_value_mg_L, maxgap = 2),
         rainwater_m3 = (ppt_mm_bret/1000)*loch_ws_size_m2, # over the entire watershed that week
         deposition_value_kg_m3 = deposition_value_mg_L / 1e+3, # Convert  concentration from mg/L to kg/m³
         deposition_kg = rainwater_m3 * deposition_value_kg_m3, #kg across the time period (a week in this case)
         deposition_kg_per_ha = deposition_kg/660) %>%
  as.data.frame() %>%
  mutate(month = case_when(weekofyear %in% c(23,24,25,26) ~ "June",
                           weekofyear %in% c(27,28,29,30) ~ "July",
                           weekofyear %in% c(31,32,33,34) ~ "August")) %>%
  filter(month %in% c("June","July","August")) %>%
  group_by(waterYear, month, deposition_name) %>%
  summarize(deposition_kg_per_ha_impute = sum(deposition_kg_per_ha, na.rm=TRUE))


#Summer dep (June-Aug)
summer_dep <- NADP %>%
  select(waterYear, weekofyear, ppt_mm_bret, Ca_mgL, Mg_mgL, K_mgL, NH4_mgL_bret, NO3_mgL_bret, IN_mgL_bret, SO4_mgL, Na_mgL) %>%
  mutate(cations_mgL = Ca_mgL + Mg_mgL + Na_mgL + K_mgL) %>%
  select(-c(Ca_mgL, Mg_mgL, Na_mgL, K_mgL)) %>%
  pivot_longer(-c(1:3),
               names_to = "deposition_name",
               values_to = "deposition_value_mg_L") %>%
  # filter(deposition_name=="NO3_mgL") %>%
  distinct(deposition_name, waterYear, weekofyear, .keep_all = TRUE) %>%
  arrange(deposition_name,waterYear, weekofyear) %>%
  as_tsibble(., key = c(deposition_name,waterYear), index = weekofyear) %>% #time series tibble
  fill_gaps() %>%
  #impute missing data
  mutate(deposition_name = case_match(
    deposition_name,
    "cations_mgL" ~ "cations",
    # "inorganicN_mgL" ~ "inorganic N",
    "NH4_mgL_bret" ~ "NH4-N",
    "NO3_mgL_bret" ~ "NO3-N",
    "SO4_mgL" ~ "SO4-S",
    "IN_mgL_bret" ~ "Inorganic N",
    .default = deposition_name
  )) %>%
  mutate(deposition_value_mg_L = case_when(deposition_name == "NO3-N" ~ deposition_value_mg_L * 0.226,
                                deposition_name == "NH4-N" ~ deposition_value_mg_L * 0.777,
                                deposition_name == "SO4-S" ~ deposition_value_mg_L * 0.333,
                                TRUE ~ deposition_value_mg_L)) %>%
  mutate(ppt_mm_bret = imputeTS::na_interpolation(ppt_mm_bret, maxgap = 2),
         deposition_value_mg_L = imputeTS::na_interpolation(deposition_value_mg_L, maxgap = 2),
         rainwater_m3 = (ppt_mm_bret/1000)*loch_ws_size_m2, # over the entire watershed that week
         deposition_value_kg_m3 = deposition_value_mg_L / 1e+3, # Convert  concentration from mg/L to kg/m³
         deposition_kg = rainwater_m3 * deposition_value_kg_m3, #kg across the time period (a week in this case)
         deposition_kg_per_ha = deposition_kg/660) %>%
  as.data.frame() %>%
  mutate(month = case_when(weekofyear %in% c(23,24,25,26) ~ "June",
                           weekofyear %in% c(27,28,29,30) ~ "July",
                           weekofyear %in% c(31,32,33,34) ~ "August")) %>%
  filter(month %in% c("June","July","August")) %>%
  mutate(season="summer") %>%
  group_by(waterYear, season, deposition_name) %>%
  summarize(deposition_kg_per_ha_impute = sum(deposition_kg_per_ha, na.rm=TRUE))



```

## Plots

### Annual deposition rates 1991-2022
```{r, echo=FALSE}

# Peak at the variability 

annual_dep %>%
  filter(waterYear >= 1991 & waterYear < 2023) %>%
  ggplot(aes(x=waterYear, y=deposition_kg_per_ha_impute, color=deposition_name))+
  geom_point()+
  geom_line()+
  facet_wrap(~deposition_name, scales="free_y")+
  theme_few(base_size = 22)
```

### Monthly deposition rates (1984-2022)
```{r, echo=FALSE, width="100%", fig.height=unit(16,"cm")}


monthly_dep %>%
  filter(waterYear >= 1984 & waterYear <= 2022) %>%
  mutate(month = factor(month, levels=c("June","July","August"))) %>%
  ggplot(aes(x=waterYear, y=deposition_kg_per_ha_impute, color=month))+
  geom_point()+
  geom_line()+
  facet_wrap(deposition_name~month, scales="free_y", ncol=3)+
  theme_few(base_size = 22)
```

### Summer deposition rates (1984-2022)
```{r, echo=FALSE}

summer_dep %>%
  filter(waterYear >= 1984 & waterYear <= 2022) %>%
  ggplot(aes(x=waterYear, y=deposition_kg_per_ha_impute, color=deposition_name))+
  geom_point()+
  geom_line()+
  facet_wrap(~deposition_name, scales="free_y")+
  theme_few(base_size = 22)

```

### Test for annual deposition trends

Demo of what's happening under the hood -- run the M-K test, and if significant extract the Sens slope and intercept.

```{r, echo=FALSE}

#Here is what's happening under the hood
N_dep_annual <- annual_dep %>%
  filter(deposition_name=="Inorganic N") %>%
  filter(waterYear > 1984 & waterYear < 2024)
# test the data for normality
hist(N_dep_annual$deposition_kg_per_ha_impute, breaks = 30)
# but let's use a Shapiro test to check for real
shapiro.test(N_dep_annual$deposition_kg_per_ha_impute)  
#P-value is below .05, therefore we reject the null hypothesis of a normal distrubution.

# We can turned to the Mann-Kendall trend test, which is non-parametric, not requiring a normal dist
# this only tells us if there is a significant trend, not the estimated slope
mk_results <-trend::mk.test(N_dep_annual$deposition_kg_per_ha_impute)
print(mk_results) # significant

# for a slope estimate we uses the Sen's slope function
sens_result <-trend::sens.slope(N_dep_annual$deposition_kg_per_ha_impute)
print(sens_result)
slope_est <-as.numeric(sens_result[[1]]) # slope estimate

# the trend package only gives us slope, not a y-int estimate for plotting
# we'll use the zyp package for that
y_int_results <-zyp::zyp.sen(deposition_kg_per_ha_impute ~ waterYear, N_dep_annual) # y comes first!!
print(y_int_results) # inspect
y_int <-y_int_results$coefficients[[1]] # pull out y-int estimate for ploting

ggplot(N_dep_annual, aes(y = deposition_kg_per_ha_impute, x = waterYear)) +
  geom_point(col = "red")+
  # use info from our MK and Sen's test to graph the trend line
  geom_abline(intercept = y_int, slope = slope_est, color="blue", size=1.2) +
  labs(title="Annual N deposition MK Trend", 
       y="Inorganic N (kg/ha)", x="Water Year")
```


```{r, echo=FALSE, include=FALSE}
#The sens slope p-value matches M-K so I built a workshop that just uses the sens slope and then the zyp package to get a y-intercept
dep_trends_nested <- annual_dep %>%
  rename(value=deposition_kg_per_ha_impute) %>%
  filter(waterYear > 1984 & waterYear < 2024) %>%
  group_by(deposition_name) %>%
  nest() %>%
  mutate(
    sens = map(data, map_sens),
    sens_sum = map(sens, broom::glance),
    slope = map(sens, sens_slope),
    zyp_mod = map(data, map_zyp),
    intercept = map(zyp_mod, sens_intercept)
  )


## Un-nest 
dep_trends_unnested = unnest(dep_trends_nested, c(sens_sum, slope, deposition_name)) %>%
  mutate(
    trend = case_when(
      p.value <= 0.05 & slope >= 0 ~ 'increasing',
      p.value <= 0.05 & slope <= 0 ~ 'decreasing',
      p.value > 0.05 ~ 'no trend'
    ),
    trend = factor(trend,
                       levels = c('no trend',
                                  'increasing',
                                  'decreasing'))
  )

#Extract all covariates for modeling...
print(dep_trends_slopes_wide <- dep_trends_unnested %>%
  select(deposition_name, slope, p.value, trend) %>%
  unnest(c(slope)))
```