---
title: "Loch outlet flux analysis"
author: "Bella Oleksy"
date: "2024-11-19"
output: html_document
---


# Some notes at the top about data processing

<b>Flux estimates--</b>
For each day, I have a solute concentration and a discharge value
solute = mg/L;
discharge = m3/s

Concentrations are interpolated between weeks as in Mast et al. 2014 ES&T

- [mg/L] * [m3/s] * [1000L/m3] * [86400s/day] = units of [kg/week]
- Divide by the area of the watershed (660ha) and you get [kg/ha/week]
- From there I summed across all weeks to get all annual yields (kg/ha/yr)

```{r setup, include=FALSE, echo=FALSE}
#source scripts
source("scripts/00_functions.R")
source("scripts/00_libraries.R")
source("scripts/01.1_LochO_data_munging.R")
# source("scripts/01.2_Bear_Lake_SNOTEL.R")
# source("scripts/01.3_NADP_CO98.R")
knitr::opts_chunk$set(fig.width=unit(12,"cm"), fig.height=unit(8,"cm"))
```

### Missing data:
Missing data are sprinkled throughout, but basically a non-issue after 1991. I interpolated between weeks with missing data where necessary but this is extremely rare after 1991. We do not have data past 2023 in the database yet.
```{r, echo=FALSE, include=FALSE}
vis_miss(outlet_raw_weekly%>%
           select(waterYear, NO3_calc)) 

```

I linearly interpolated missing Loch outlet chemistry values between sampling trips up with up to 30 days of missing data. Beyond that, I did not interpolate and those are in the dataset as NAs.

```{r, echo=FALSE, include=FALSE}
outlet_weekly_flux %>%
  group_by(chem_name, waterYear) %>%
  summarize(n=n(),
         n_NAs=sum(is.na(chem_value)),
         perc_missing_days=(n_NAs/365)*100) %>%# calculate percentage of missing observations
  ggplot(aes(x=waterYear, y=perc_missing_days, color=chem_name))+
  geom_point()+
  facet_wrap(~chem_name)+
  labs(title="% missing data annually")

```

There's still some 2019 NO3 data missing that I will ask MJ's help to track down.

```{r, echo=FALSE, include=FALSE}
outlet_weekly_flux %>%
  mutate(month = month(DATE)) %>%
  filter(month %in% c(6,7,8)) %>%
  group_by(chem_name, waterYear) %>%
  summarize(n=n(),
         n_NAs=sum(is.na(chem_value)),
         perc_missing_days=(n_NAs/90)*100) %>%# calculate percentage of missing observations
  ggplot(aes(x=waterYear, y=perc_missing_days, color=chem_name))+
  geom_point()+
  facet_wrap(~chem_name)+
  labs(title="% missing data June-August")
```

Summer data are robust 1991-2022



### Test for annual export trends

Mann-Kendell test with Sens Slope

```{r, echo=FALSE}

##ANNUAL
annual_export_trends_nested <- annual_flux %>%
  rename(value=annual_flux_kg_per_ha) %>%
  filter(waterYear >= 1991 & waterYear <= 2022) %>%
  group_by(chem_name) %>%
  nest() %>%
  mutate(
    sens = map(data, map_sens),
    sens_sum = map(sens, broom::glance),
    slope = map(sens, sens_slope),
    zyp_mod = map(data, map_zyp),
    intercept = map(zyp_mod, sens_intercept)
  )


## Un-nest 
annual_export_trends_unnested = unnest(annual_export_trends_nested, c(sens_sum, slope, chem_name)) %>%
  mutate(
    trend = case_when(
      p.value <= 0.05 & slope >= 0 ~ 'increasing',
      p.value <= 0.05 & slope <= 0 ~ 'decreasing',
      p.value > 0.05 ~ 'no trend'
    ),
    trend = factor(trend,
                       levels = c('no trend',
                                  'increasing',
                                  'decreasing'))
  )

#Extract all covariates for modeling...
annual_export_trends <- annual_export_trends_unnested %>%
  select(chem_name, slope, p.value, trend, intercept) %>%
  unnest(c(slope)) %>%
  mutate(time_frame = "annual")%>%
  mutate(intercept = as.numeric(intercept))


##MONTHLY
monthly_export <- bind_rows( june_flux %>%
  select(-june_flux_kg) %>%
  mutate(month="June") %>%
  rename(value=june_flux_kg_per_ha),
  july_flux %>%
  select(-july_flux_kg) %>%
  mutate(month="July") %>%
  rename(value=july_flux_kg_per_ha),
  august_flux %>%
  select(-august_flux_kg) %>%
  mutate(month="August") %>%
  rename(value=august_flux_kg_per_ha))


monthly_export_trends_nested <- monthly_export %>%
  filter(waterYear > 1984 & waterYear < 2023) %>%
  group_by(chem_name, month) %>%
  nest() %>%
  mutate(
    sens = map(data, map_sens),
    sens_sum = map(sens, broom::glance),
    slope = map(sens, sens_slope),
    zyp_mod = map(data, map_zyp),
    intercept = map(zyp_mod, sens_intercept)
  )


## Un-nest 
monthly_export_trends_unnested = unnest(monthly_export_trends_nested, c(sens_sum, slope, chem_name, month)) %>%
  mutate(
    trend = case_when(
      p.value <= 0.05 & slope >= 0 ~ 'increasing',
      p.value <= 0.05 & slope <= 0 ~ 'decreasing',
      p.value > 0.05 ~ 'no trend'
    ),
    trend = factor(trend,
                       levels = c('no trend',
                                  'increasing',
                                  'decreasing'))
  )

#Extract all covariates for modeling...
monthly_export_trends <- monthly_export_trends_unnested %>%
  select(chem_name, month, slope, p.value, trend, intercept) %>%
  unnest(c(slope)) %>%
  rename(time_frame=month)%>%
  mutate(intercept = as.numeric(intercept))


##SUMMER

summer_export_trends_nested <- summer_flux %>%
  rename(value=summer_flux_kg_per_ha) %>%
  filter(waterYear > 1984 & waterYear <= 2022) %>%
  group_by(chem_name) %>%
  nest() %>%
  mutate(
    sens = map(data, map_sens),
    sens_sum = map(sens, broom::glance),
    slope = map(sens, sens_slope),
    zyp_mod = map(data, map_zyp),
    intercept = map(zyp_mod, sens_intercept)
  )


## Un-nest 
summer_export_trends_unnested = unnest(summer_export_trends_nested, c(sens_sum, slope, chem_name)) %>%
  mutate(
    trend = case_when(
      p.value <= 0.05 & slope >= 0 ~ 'increasing',
      p.value <= 0.05 & slope <= 0 ~ 'decreasing',
      p.value > 0.05 ~ 'no trend'
    ),
    trend = factor(trend,
                       levels = c('no trend',
                                  'increasing',
                                  'decreasing'))
  )

#Extract all covariates for modeling...
summer_export_trends <- summer_export_trends_unnested %>%
  select(chem_name, slope, p.value, trend, intercept) %>%
  unnest(c(slope)) %>%
  mutate(time_frame = "summer") %>%
  mutate(intercept = as.numeric(intercept))


all_export_trends <- bind_rows(annual_export_trends,
                            summer_export_trends,
                            monthly_export_trends) %>%
  mutate(slope = round(slope, 3),
         p.value = round(p.value,4),
         p.value = case_when(p.value < 0.001 ~ "<0.001",
                             TRUE ~ as.character(p.value))) 

all_export_trends_hux <- 
  hux(all_export_trends) %>% 
  # add_colnames() %>% 
  set_bold(row = 1, col = everywhere, value = TRUE) %>% 
  set_all_borders(TRUE) 

theme_plain(all_export_trends_hux)
quick_docx(all_export_trends_hux, file = 'figures/Table2_outlet_export_trends.docx')

```

## Plots

### Annual fluxes 1991-2022

There is only a trend line on the plot if p < 0.05.

```{r, echo=FALSE, warning=FALSE}

annual_flux %>%
  filter(waterYear >= 1991 & waterYear <= 2024) %>%
  ggplot(aes(x=waterYear, y=annual_flux_kg_per_ha, fill=chem_name, group=chem_name))+
  geom_point(size=4, shape=21)+ 
  facet_wrap(~chem_name, scales="free_y")+
  theme_few(base_size = 24)+
  geom_abline(data = annual_export_trends %>% filter(!trend=="no trend"), aes(intercept = intercept, slope = slope),
              col = "black",
              linewidth=1.5) +
  labs(y="Flux (kg/ha/year)",
       x="Year",
       title="Annual") +
  theme(legend.position = "none")+
  scale_color_manual(values=chemCols,
                     "Fit")+
  scale_fill_manual(values=chemCols,
                    name="Observed")
```

### Monthly fluxes 1984-2022

There is only a trend line on the plot if p < 0.05.

```{r, echo=FALSE, warning=FALSE}

monthly_export %>%
  filter(waterYear > 1984 & waterYear < 2023) %>%
  ggplot(aes(x=waterYear, y=value, fill=month, group=month))+
  geom_point(size=4, shape=21)+ 
  facet_wrap(~chem_name, scales="free_y")+
  theme_few(base_size = 24)+
  geom_abline(data = monthly_export_trends %>% rename(month=time_frame) %>%filter(!trend=="no trend"), aes(intercept = intercept, slope = slope, color=month),
              linewidth=1.5) +
  labs(y="Flux (kg/ha/year)",
       x="Year",
       title="Monthly") +
  scale_color_manual(values=monthCols,
                     "Fit")+
  scale_fill_manual(values=monthCols,
                    name="Observed")
```

### Summer fluxes 1984-2022

There is only a trend line on the plot if p < 0.05.

```{r, echo=FALSE, warning=FALSE}

summer_flux %>%
  filter(waterYear > 1984 & waterYear < 2023) %>%
  ggplot(aes(x=waterYear, y=summer_flux_kg_per_ha, fill=chem_name, group=chem_name))+
  geom_point(size=4, shape=21)+ 
  facet_wrap(~chem_name, scales="free_y")+
  theme_few(base_size = 24)+
  geom_abline(data = summer_export_trends %>%filter(!trend=="no trend"), aes(intercept = intercept, slope = slope, color=chem_name),
              linewidth=1.5) +
  labs(y="Flux (kg/ha/year)",
       x="Year",
       title="Summer (JJA)") +
  scale_color_manual(values=chemCols,
                     "Fit")+
  scale_fill_manual(values=chemCols,
                    name="Observed")
```
