---
title: "outlet_trend_analysis"
author: "Bella Oleksy"
date: "2024-11-19"
output: html_document
---


# Some notes at the top about data processing

<b>Flux estimates--</b>
For each day, I have a solute concentration and a discharge value
solute = mg/L;
discharge = m3/s

Concentrations are interpolated between weeks as in Mast et al. 2014 ES&T

- [mg/L] * [m3/s] * [1000L/m3] * [86400s/day] = units of [kg/week]
- Divide by the area of the watershed (660ha) and you get [kg/ha/week]
- From there I summed across all weeks to get all annual yields (kg/ha/yr)

```{r setup, include=FALSE, echo=FALSE}
#source scripts
source("scripts/00_functions.R")
source("scripts/00_libraries.R")
source("scripts/01.1_LochO_data_munging.R")
# source("scripts/01.2_Bear_Lake_SNOTEL.R")
# source("scripts/01.3_NADP_CO98.R")
knitr::opts_chunk$set(fig.width=unit(12,"cm"), fig.height=unit(8,"cm"))
```

### Missing data:
Missing data are sprinkled throughout, but basically a non-issue after 1991. I interpolated between weeks with missing data where necessary but this is extremely rare after 1991. We do not have data past 2023 in the database yet.
```{r, echo=FALSE, include=FALSE}
vis_miss(outlet_raw_weekly%>%
           select(waterYear, NO3_calc)) 

```

I linearly interpolated missing Loch outlet chemistry values between sampling trips up with up to 30 days of missing data. Beyond that, I did not interpolate and those are in the dataset as NAs.

```{r, echo=FALSE, include=FALSE}
outlet_weekly_flux %>%
  group_by(chem_name, waterYear) %>%
  summarize(n=n(),
         n_NAs=sum(is.na(chem_value)),
         perc_missing_days=(n_NAs/365)*100) %>%# calculate percentage of missing observations
  ggplot(aes(x=waterYear, y=perc_missing_days, color=chem_name))+
  geom_point()+
  facet_wrap(~chem_name)+
  labs(title="% missing data annually")

```

There's still some 2019 NO3 data missing that I will ask MJ's help to track down.

```{r, echo=FALSE, include=FALSE}
outlet_weekly_flux %>%
  mutate(month = month(DATE)) %>%
  filter(month %in% c(6,7,8)) %>%
  group_by(chem_name, waterYear) %>%
  summarize(n=n(),
         n_NAs=sum(is.na(chem_value)),
         perc_missing_days=(n_NAs/90)*100) %>%# calculate percentage of missing observations
  ggplot(aes(x=waterYear, y=perc_missing_days, color=chem_name))+
  geom_point()+
  facet_wrap(~chem_name)+
  labs(title="% missing data June-August")
```

Summer data are robust 1991-2022

## Plots

### Annual fluxes 1991-2022
```{r, echo=FALSE, width="100%"}
# Peak at the variability 
annual_flux %>%
  filter(waterYear >= 1991 & waterYear <= 2024) %>%
  ggplot(aes(x=waterYear, y=annual_flux_kg_per_ha, color=chem_name))+
  geom_point()+
  geom_line()+
  facet_wrap(~chem_name, scales="free_y")+
  theme_few(base_size = 24)
```

### Monthly fluxes 1984-2022
```{r, echo=FALSE, width="100%", fig.height=unit(16,"cm")}

# Peak at the variability

june_flux %>%
  filter(waterYear >= 1984 & waterYear <= 2024) %>%
  ggplot(aes(x=waterYear, y=june_flux_kg_per_ha, color=chem_name))+
  geom_point()+
  geom_line()+
  facet_wrap(~chem_name, scales="free_y")+
  theme_few(base_size = 24)+
  labs(title="June")

july_flux %>%
  filter(waterYear >= 1991 & waterYear <= 2024) %>%
  ggplot(aes(x=waterYear, y=july_flux_kg_per_ha, color=chem_name))+
  geom_point()+
  geom_line()+
  facet_wrap(~chem_name, scales="free_y")+
  theme_few(base_size = 24)+
  labs(title="July")

august_flux %>%
  filter(waterYear >= 1991 & waterYear <= 2024) %>%
  ggplot(aes(x=waterYear, y=august_flux_kg_per_ha, color=chem_name))+
  geom_point()+
  geom_line()+
  facet_wrap(~chem_name, scales="free_y")+
  theme_few(base_size = 24)+
  labs(title="August")
```

### Summer fluxes 1984-2022
```{r, echo=FALSE, width="100%", fig.height=unit(16,"cm")}

# Peak at the variability

summer_flux %>%
  filter(waterYear >= 1984 & waterYear <= 2022) %>%
  ggplot(aes(x=waterYear, y=summer_flux_kg_per_ha, color=chem_name))+
  geom_point()+
  geom_line()+
  facet_wrap(~chem_name, scales="free_y")+
  theme_few(base_size = 24)+
  labs(title="Summer (JJA)")
```
