---
title: "outlet_trend_analysis"
author: "Bella Oleksy"
date: "2024-11-19"
output: html_document
---


# Some notes at the top about data processing

<b>Flux estimates--</b>
- The numbers don't match what appears in Heath & Baron 2013 nor other papers like Mast et al 2014 ES&T
- For example, NO3 annual yield should be in the realm of 1-4 kg/ha/yr
- Here is how I did my calculations:

For each week, I have a solute concentration and a discharge value
solute = mg/L
discharge = m3/s

[mg/L] * [m3/s] * [1000L/m3] * [86400s/day]

```{r setup, include=FALSE, echo=FALSE}
#source scripts
source("scripts/00_functions.R")
source("scripts/00_libraries.R")
source("scripts/01.1_LochO_data_munging.R")
# source("scripts/01.2_Bear_Lake_SNOTEL.R")
# source("scripts/01.3_NADP_CO98.R")
knitr::opts_chunk$set(fig.width=unit(12,"cm"), fig.height=unit(8,"cm"))
```

## Missing data:
Missing data are sprinkled throughout, but basically a non-issue. I interpolated between weeks with missing data where necessary but this is extremely rare after 1991. We do not have data past 2023 in the database yet.
```{r, echo=FALSE}
vis_miss(outlet_raw_weekly%>%
           select(waterYear, NO3_calc)) 

```

```{r, echo=FALSE, include=FALSE}

annual_LochO_flux <- outlet_weekly_flux %>%
  filter(waterYear >= 1991 & waterYear <= 2022) %>%
  group_by(waterYear, chem_name) %>%
  summarize(flux_kg_per_ha_year = sum(flux_kg_per_ha, na.rm=TRUE))


monthly_LochO_flux <- outlet_weekly_flux %>%
  mutate(month = case_when(weekofyear %in% c(23,24,25,26) ~ "June",
                           weekofyear %in% c(27,28,29,30) ~ "July",
                           weekofyear %in% c(31,32,33,34) ~ "August")) %>%
  filter(month %in% c("June","July","August")) %>%
  group_by(waterYear, month, chem_name) %>%
  summarize(flux_kg_per_ha_year = sum(flux_kg_per_ha, na.rm=TRUE))


summer_LochO_flux <- outlet_weekly_flux %>%
  mutate(month = case_when(weekofyear %in% c(23,24,25,26) ~ "June",
                           weekofyear %in% c(27,28,29,30) ~ "July",
                           weekofyear %in% c(31,32,33,34) ~ "August")) %>%
  filter(month %in% c("June","July","August")) %>%
  mutate(season="summer") %>%
  group_by(waterYear, season, chem_name) %>%
  summarize(flux_kg_per_ha_year = sum(flux_kg_per_ha, na.rm=TRUE))

```

## Plots

### Weekly fluxes 1991-2024
```{r}
outlet_weekly_flux %>%
  filter(waterYear >= 1991 & waterYear <= 2022) %>%
  ggplot(aes(x=waterYear, y=flux_kg_per_ha, color=chem_name))+
  geom_point()+
  geom_line()+
  facet_wrap(~chem_name, scales="free_y")+
  theme_few(base_size = 24)+
  labs(y="kg per ha per week")

```

### Annual fluxes 1991-2024
```{r, echo=FALSE, width="100%"}

# Peak at the variability 

annual_LochO_flux %>%
  filter(waterYear >= 1991 & waterYear <= 2024) %>%
  ggplot(aes(x=waterYear, y=flux_kg_per_ha_year, color=chem_name))+
  geom_point()+
  geom_line()+
  facet_wrap(~chem_name, scales="free_y")+
  theme_few(base_size = 24)
```

### Monthly fluxes 1984-2024
```{r, echo=FALSE, width="100%", fig.height=unit(16,"cm")}

# Peak at the variability 

monthly_LochO_flux %>%
  filter(waterYear >= 1984 & waterYear <= 2024) %>%
  mutate(month = factor(month, levels=c("June","July","August"))) %>%
  ggplot(aes(x=waterYear, y=flux_kg_per_ha_year, color=chem_name))+
  geom_point()+
  geom_line()+
  facet_wrap(~chem_name, scales="free_y")+
  facet_wrap(chem_name~month, scales="free_y", ncol=3)+
  theme_few(base_size = 24)
```

