---
title: "Loch outlet flux analysis"
author: "Bella Oleksy"
date: "2024-11-19"
output: html_document
---


# Some notes at the top about data processing

<b>Flux estimates--</b>
For each day, I have a solute concentration and a discharge value
solute = mg/L;
discharge = m3/s

Concentrations are interpolated between weeks as in Mast et al. 2014 ES&T

- [mg/L] * [m3/s] * [1000L/m3] * [86400s/day] = units of [kg/week]
- Divide by the area of the watershed (660ha) and you get [kg/ha/week]
- From there I summed across all weeks to get all annual yields (kg/ha/yr)

```{r setup, include=FALSE, echo=FALSE}
#source scripts
source("scripts/00_functions.R")
source("scripts/00_libraries.R")
source("scripts/01.1_LochO_data_munging.R")
# source("scripts/01.2_Bear_Lake_SNOTEL.R")
# source("scripts/01.3_NADP_CO98.R")
knitr::opts_chunk$set(fig.width=unit(12,"cm"), fig.height=unit(8,"cm"))
```

### Missing data:
Missing data are sprinkled throughout, but basically a non-issue after 1991. I interpolated between weeks with missing data where necessary but this is extremely rare after 1991. We do not have data past 2023 in the database yet.
```{r, echo=FALSE, include=FALSE}
# vis_miss(outlet_raw_weekly%>%
#            select(waterYear, no3_mgl)) 

```

I linearly interpolated missing Loch outlet chemistry values between sampling trips up with up to 30 days of missing data. Beyond that, I did not interpolate and those are in the dataset as NAs.

```{r, echo=FALSE, include=FALSE}
# outlet_weekly_flux %>%
#   group_by(chem_name, waterYear) %>%
#   summarize(n=n(),
#          n_NAs=sum(is.na(chem_value)),
#          perc_missing_days=(n_NAs/365)*100) %>%# calculate percentage of missing observations
#   ggplot(aes(x=waterYear, y=perc_missing_days, color=chem_name))+
#   geom_point()+
#   facet_wrap(~chem_name)+
#   labs(title="% missing data annually")

```

There's still some 2019 NO3 data missing that I will ask MJ's help to track down.

```{r, echo=FALSE, include=FALSE}
outlet_daily_flux %>%
  mutate(month = month(date)) %>%
  filter(month %in% c(6,7,8)) %>%
  group_by(chem_name, waterYear) %>%
  summarize(n=n(),
         n_NAs=sum(is.na(chem_value)),
         perc_missing_days=(n_NAs/90)*100) %>%# calculate percentage of missing observations
  ggplot(aes(x=waterYear, y=perc_missing_days, color=chem_name))+
  geom_point()+
  facet_wrap(~chem_name)+
  labs(title="% missing data June-August")
```

Summer data are robust 1991-2022



### Test for annual export trends

Mann-Kendell test with Sens Slope

```{r, echo=FALSE}

##ANNUAL
annual_export_trends_nested <- annual_flux %>%
  rename(value=annual_flux_kg_per_ha) %>%
  filter(waterYear >= 1991 & waterYear <= 2022) %>%
  group_by(chem_name) %>%
  nest() %>%
  mutate(
    sens = map(data, map_sens),
    sens_sum = map(sens, broom::glance),
    slope = map(sens, sens_slope),
    zyp_mod = map(data, map_zyp),
    intercept = map(zyp_mod, sens_intercept)
  )


## Un-nest 
annual_export_trends_unnested = unnest(annual_export_trends_nested, c(sens_sum, slope, chem_name)) %>%
  mutate(
    trend = case_when(
      p.value <= 0.05 & slope >= 0 ~ 'increasing',
      p.value <= 0.05 & slope <= 0 ~ 'decreasing',
      p.value > 0.05 ~ 'no trend'
    ),
    trend = factor(trend,
                       levels = c('no trend',
                                  'increasing',
                                  'decreasing'))
  )

#Extract all covariates for modeling...
annual_export_trends <- annual_export_trends_unnested %>%
  select(chem_name, slope, p.value, trend, intercept) %>%
  unnest(c(slope)) %>%
  mutate(time_frame = "annual")%>%
  mutate(intercept = as.numeric(intercept))


##MONTHLY
monthly_flux <- bind_rows( june_flux %>%
  select(-june_flux_kg) %>%
  mutate(month="June") %>%
  rename(value=june_flux_kg_per_ha),
  july_flux %>%
  select(-july_flux_kg) %>%
  mutate(month="July") %>%
  rename(value=july_flux_kg_per_ha),
  august_flux %>%
  select(-august_flux_kg) %>%
  mutate(month="August") %>%
  rename(value=august_flux_kg_per_ha))


monthly_flux_trends_nested <- monthly_flux %>%
  filter(waterYear > 1984 & waterYear < 2023) %>%
  filter(!waterYear == 1989) %>% #more than half the data missing this year
  group_by(chem_name, month) %>%
  nest() %>%
  mutate(
    sens = map(data, map_sens),
    sens_sum = map(sens, broom::glance),
    slope = map(sens, sens_slope),
    zyp_mod = map(data, map_zyp),
    intercept = map(zyp_mod, sens_intercept)
  )


## Un-nest 
monthly_flux_trends_unnested = unnest(monthly_flux_trends_nested, c(sens_sum, slope, chem_name, month)) %>%
  mutate(
    trend = case_when(
      p.value <= 0.05 & slope >= 0 ~ 'increasing',
      p.value <= 0.05 & slope <= 0 ~ 'decreasing',
      p.value > 0.05 ~ 'no trend'
    ),
    trend = factor(trend,
                       levels = c('no trend',
                                  'increasing',
                                  'decreasing'))
  )

#Extract all covariates for modeling...
monthly_flux_trends <- monthly_flux_trends_unnested %>%
  select(chem_name, month, slope, p.value, trend, intercept) %>%
  unnest(c(slope)) %>%
  rename(time_frame=month)%>%
  mutate(intercept = as.numeric(intercept))


##SUMMER

summer_export_trends_nested <- summer_flux %>%
  rename(value=summer_flux_kg_per_ha) %>%
  filter(waterYear > 1984 & waterYear < 2023) %>%
  filter(!waterYear == 1989) %>% #more than half the data missing this year
  group_by(chem_name) %>%
  nest() %>%
  mutate(
    sens = map(data, map_sens),
    sens_sum = map(sens, broom::glance),
    slope = map(sens, sens_slope),
    zyp_mod = map(data, map_zyp),
    intercept = map(zyp_mod, sens_intercept)
  )


## Un-nest 
summer_export_trends_unnested = unnest(summer_export_trends_nested, c(sens_sum, slope, chem_name)) %>%
  mutate(
    trend = case_when(
      p.value <= 0.05 & slope >= 0 ~ 'increasing',
      p.value <= 0.05 & slope <= 0 ~ 'decreasing',
      p.value > 0.05 ~ 'no trend'
    ),
    trend = factor(trend,
                       levels = c('no trend',
                                  'increasing',
                                  'decreasing'))
  )

#Extract all covariates for modeling...
summer_export_trends <- summer_export_trends_unnested %>%
  select(chem_name, slope, p.value, trend, intercept) %>%
  unnest(c(slope)) %>%
  mutate(time_frame = "summer") %>%
  mutate(intercept = as.numeric(intercept))


all_export_trends <- bind_rows(annual_export_trends,
                            summer_export_trends,
                            monthly_flux_trends) %>%
  mutate(slope = round(slope, 3),
         p.value = round(p.value,4),
         p.value = case_when(p.value < 0.001 ~ "<0.001",
                             TRUE ~ as.character(p.value))) 

all_export_trends_hux <- 
  hux(all_export_trends) %>% 
  # add_colnames() %>% 
  set_bold(row = 1, col = everywhere, value = TRUE) %>% 
  set_all_borders(TRUE) 

theme_plain(all_export_trends_hux)
quick_docx(all_export_trends_hux, file = 'figures/Table2_outlet_export_trends.docx')

```

## Plots

### Annual fluxes 1991-2022

There is only a trend line on the plot if p < 0.05.

```{r, echo=FALSE, warning=FALSE}

annual_flux %>%
  filter(waterYear >= 1991 & waterYear <= 2024) %>%
  ggplot(aes(x=waterYear, y=annual_flux_kg_per_ha, fill=chem_name, group=chem_name))+
  geom_point(size=2, shape=21)+ 
  facet_wrap(~chem_name, scales="free_y")+
  theme_few(base_size = 12)+
  geom_abline(data = annual_export_trends %>% filter(!trend=="no trend"), aes(intercept = intercept, slope = slope, color=chem_name)) +
  labs(y="Flux (kg/ha/year)",
       x="Year",
       title="Annual") +
  theme(legend.position = "none")+
  scale_color_manual(values=chemCols,
                     "Fit")+
  scale_fill_manual(values=chemCols,
                    name="Observed")
ggsave("figures/loch_export_trends_annual.png", dpi=600, width=8, height=6, units="in")
```

### Monthly fluxes 1984-2022

There is only a trend line on the plot if p < 0.05.

```{r, echo=FALSE, warning=FALSE}

monthly_flux %>%
  filter(waterYear > 1984 & waterYear < 2023) %>%
  filter(!waterYear == 1989) %>% #more than half the data missing this year
  ggplot(aes(x=waterYear, y=value, fill=month, group=month))+
  geom_point(size=2, shape=21)+ 
  facet_wrap(~chem_name, scales="free_y")+
  theme_few(base_size = 12)+
  geom_abline(data = monthly_flux_trends %>% rename(month=time_frame) %>%filter(!trend=="no trend"), aes(intercept = intercept, slope = slope, color=month)) +
  labs(y="Flux (kg/ha/year)",
       x="Year",
       title="Monthly") +
  scale_color_manual(values=monthCols,
                     "Fit")+
  scale_fill_manual(values=monthCols,
                    name="Observed")
ggsave("figures/loch_export_trends_monthly.png", dpi=600, width=8, height=6, units="in")
```

### Summer fluxes 1984-2022

There is only a trend line on the plot if p < 0.05.

```{r, echo=FALSE, warning=FALSE}

summer_flux %>%
  filter(waterYear > 1984 & waterYear < 2023) %>%
  filter(!waterYear == 1989) %>% #more than half the data missing this year
  ggplot(aes(x=waterYear, y=summer_flux_kg_per_ha, fill=chem_name, group=chem_name))+
  geom_point(size=2, shape=21)+ 
  facet_wrap(~chem_name, scales="free_y")+
  theme_few(base_size = 12)+
  geom_abline(data = summer_export_trends %>%filter(!trend=="no trend"), aes(intercept = intercept, slope = slope, color=chem_name)) +
  labs(y="Flux (kg/ha/year)",
       x="Year",
       title="Summer (JJA)") +
  scale_color_manual(values=chemCols,
                     "Fit")+
  scale_fill_manual(values=chemCols,
                    name="Observed")
ggsave("figures/loch_export_trends_summer.png", dpi=600, width=8, height=6, units="in")
```

# Look at hydrologic season definitions
```{r}

CQ_seasons <- outlet_daily_flux %>%
  left_join(., percentile_days, by="waterYear") %>%
  mutate(wy_doy = hydro.day(date)) %>%
  mutate(hydro_season = case_when(date > day_80th  ~ "fall baseflow",
                                  date < day_20th ~ "winter baseflow",
                                  date >= day_20th & date < day_50th ~ "rising limb",
                                  date >= day_50th & date <= day_80th ~ "falling limb")) 
head(CQ_seasons)

# Visualize missingness first, which years/solutes are sus?
CQ_seasons %>%
  group_by(chem_name, waterYear, hydro_season) %>%
  summarize(n=n(),
         n_NAs=sum(is.na(chem_value)),
         perc_missing_days=(n_NAs/365)*100) %>%# calculate percentage of missing observations
  ggplot(aes(x=waterYear, y=perc_missing_days, color=chem_name))+
  geom_point()+
  facet_wrap(hydro_season~chem_name)+
  labs(title="% missing data annually")

hydro_season_flux <- CQ_seasons %>%
    filter(!chem_name %in% c("temp_c","tdn_mgl","po4_mgl")) %>%
  group_by(chem_name, waterYear, hydro_season) %>%
  mutate(n=n(),
         n_NAs=sum(is.na(chem_value)),
         perc_missing_days=(n_NAs/365)*100) %>%
  filter(!perc_missing_days > 20) %>% #drop year*chem*hydroseason combos with a lot of NAs
  select(-c(n, n_NAs, perc_missing_days)) %>%
  group_by(waterYear, chem_name, hydro_season) %>%
  summarise(
    # Calculate annual flux over that time period
    # annual_flux_kg = sum(daily_flux_kg, na.rm = TRUE)) %>%
    annual_flux_kg = sum(daily_flux_kg, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    # Calculate annual flux per hectare
    annual_flux_kg_per_ha = annual_flux_kg / 660)



##hydro seasons
hydroseason_export_trends_nested <- hydro_season_flux %>%
  rename(value=annual_flux_kg_per_ha) %>%
  # filter(waterYear >= 1991 & waterYear <= 2022) %>%
  group_by(hydro_season, chem_name) %>%
  nest() %>%
  mutate(
    sens = map(data, map_sens),
    sens_sum = map(sens, broom::glance),
    slope = map(sens, sens_slope),
    zyp_mod = map(data, map_zyp),
    intercept = map(zyp_mod, sens_intercept)
  )


## Un-nest 
hydroseason_export_trends_unnested = unnest(hydroseason_export_trends_nested, c(sens_sum, slope, hydro_season, chem_name)) %>%
  mutate(
    trend = case_when(
      p.value <= 0.05 & slope >= 0 ~ 'increasing',
      p.value <= 0.05 & slope <= 0 ~ 'decreasing',
      p.value > 0.05 ~ 'no trend'
    ),
    trend = factor(trend,
                       levels = c('no trend',
                                  'increasing',
                                  'decreasing'))
  )

#Extract all covariates for modeling...
hydroseason_export_trends <- hydroseason_export_trends_unnested %>%
  select(hydro_season, chem_name, slope, p.value, trend, intercept) %>%
  unnest(c(slope)) %>%
  mutate(intercept = as.numeric(intercept))
```

## Plot hydro season fluxes
```{r}

hydro_season_flux %>%
  mutate(
    hydro_season = factor(hydro_season,
                          levels = c("rising limb",
                                     "falling limb",
                                     "fall baseflow",
                                     "winter baseflow")),
    chem_name = case_match(
      chem_name,
      "sio2_mgl" ~ "silica",
      .default = chem_name
    )
  ) %>%
  ggplot(aes(x = waterYear, y = annual_flux_kg_per_ha)) +
  
  geom_point(aes(fill = hydro_season), 
             size = 2, shape = 21, color = "black", alpha = 0.7) +
  
  geom_abline(data = hydroseason_export_trends %>%
                filter(!trend == "no trend") %>%
                mutate(chem_name = case_match(
                  chem_name,
                  "sio2_mgl" ~ "silica",
                  .default = chem_name
                )),
              aes(intercept = intercept, slope = slope, color = hydro_season),
              linewidth = 1.5,
              show.legend = FALSE) +
  
  facet_wrap(~chem_name, scales = "free_y") +
  theme_few(base_size = 12) +
  theme(legend.position = "bottom") +  # 👈 move legend
  
  labs(y = "Flux (kg/ha/year)",
       x = "Year") +
  
  scale_fill_manual(values = hydroCols, name = "Hydrologic season") +
  scale_color_manual(values = hydroCols) +
  
  guides(fill = guide_legend(nrow = 2, byrow = TRUE, override.aes = list(alpha = 1)))  


ggsave("figures/loch_export_trends_hydroseason.png", dpi=600, width=8, height=6, units="in")
# Change in NH4 as a proportion of DIN?
hydro_season_flux %>%
  ungroup() %>%
  select(waterYear, hydro_season, chem_name, annual_flux_kg_per_ha) %>%
  filter(chem_name %in% c("NH4-N","NO3-N")) %>%
  pivot_wider(names_from=chem_name,
              values_from=annual_flux_kg_per_ha) %>%
  mutate(DIN = `NH4-N`+`NO3-N`,
         NH4_DIN = `NH4-N`/DIN,
         NO3_DIN = `NO3-N`/DIN) %>%
  ggplot(aes(x=waterYear, y=NO3_DIN, color=hydro_season, fill=hydro_season, group=hydro_season))+
  geom_point(size=2, shape=21, color="black")+ 
  theme_few(base_size = 12)+
  geom_smooth(method="lm")+
  scale_color_manual(values=hydroCols,
                     "Fit")+
  scale_fill_manual(values=hydroCols,
                    name="Observed")

hydro_season_flux %>%
  ungroup() %>%
  select(waterYear, hydro_season, chem_name, annual_flux_kg_per_ha) %>%
  filter(chem_name %in% c("NH4-N","NO3-N")) %>%
  pivot_wider(names_from=chem_name,
              values_from=annual_flux_kg_per_ha) %>%
  mutate(DIN = `NH4-N`+`NO3-N`,
         NH4_DIN = `NH4-N`/DIN,
         NO3_DIN = `NO3-N`/DIN) %>%
  ggplot(aes(x=waterYear, y=NH4_DIN, color=hydro_season, fill=hydro_season, group=hydro_season))+
  geom_point(size=2, shape=21, color="black")+ 
  theme_few(base_size = 12)+
  geom_smooth(method="gam")+
  scale_color_manual(values=hydroCols,
                     "Legend")+
  scale_fill_manual(values=hydroCols,
                    name="Legend")


```
